---
title: "Lineage Overview: registering sources, transforms, and datasets"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Lineage Overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Introduction

`lineageR` builds a directed acyclic graph (DAG) that tracks where data
comes from, what transformed it, and what would break if it changed.

Four node types exist in the DAG:

| Type          | Meaning                                   |
|---------------|-------------------------------------------|
| `source`      | External origin (database, S3, API, …)   |
| `transform`   | A computation step (R script, SQL, …)    |
| `dataset`     | A table or file produced by a transform  |
| `object_type` | An ontologySpecR object type             |

## Setting up a context

```{r, eval = FALSE}
library(lineageR)
library(duckdb)

con <- dbConnect(duckdb(), ":memory:")
ctx <- lineage_context(con)
```

`lineage_context()` creates the three schema tables (`lineage_nodes`,
`lineage_edges`, `lineage_run_log`) if they do not already exist, making
initialisation idempotent.

## Registering nodes

```{r, eval = FALSE}
# Data sources
s3_id <- lin_register_source(ctx, "s3_raw_flights",
  connection_string = "s3://my-bucket/raw/flights/")

pg_id <- lin_register_source(ctx, "postgres_schedules",
  connection_string = "postgresql://host/db")

# Transform steps
ingest_id <- lin_register_transform(ctx, "ingest_flights",
  script_path = "pipelines/ingest.R",
  language    = "r")

clean_id  <- lin_register_transform(ctx, "clean_airports",
  script_path = "pipelines/clean.R",
  language    = "r")

# Datasets (tables) produced by the transforms
raw_id    <- lin_register_dataset(ctx, "flights_raw",    table_name = "flights_raw")
clean_id2 <- lin_register_dataset(ctx, "airports_clean", table_name = "airports_clean")

# Object types backed by datasets
airport_id <- lin_register_object_type(ctx, "Airport")
```

## Wiring edges

```{r, eval = FALSE}
lin_add_input(ctx,  ingest_id, s3_id)       # s3   feeds ingest
lin_add_output(ctx, ingest_id, raw_id)      # ingest produces flights_raw

lin_add_input(ctx,  clean_id,  raw_id)      # flights_raw feeds clean
lin_add_output(ctx, clean_id,  clean_id2)   # clean produces airports_clean

lin_add_backing(ctx, clean_id2, airport_id) # airports_clean backs Airport
```

Attempting to add an edge that would create a cycle is detected immediately:

```{r, eval = FALSE}
# This would make Airport an ancestor of itself — aborts with an error.
lin_add_input(ctx, s3_id, airport_id)
```

## Querying provenance

*What raw sources feed the `Airport` object type?*

```{r, eval = FALSE}
lin_provenance(ctx, airport_id)
#>          node_id  node_type           name distance last_updated_at
#> 1 <uuid>          dataset  airports_clean        1              NA
#> 2 <uuid>          transform  clean_airports      2              NA
#> 3 <uuid>          dataset    flights_raw         3              NA
#> 4 <uuid>          transform  ingest_flights      4              NA
#> 5 <uuid>          source     s3_raw_flights      5              NA

lin_provenance(ctx, airport_id, depth = 1)
#> Direct parent only: airports_clean
```

## Querying impact

*If the S3 source changes, what nodes are affected?*

```{r, eval = FALSE}
lin_impact(ctx, s3_id)
```

## Checking freshness

After recording a successful run the output datasets' `last_updated_at` is
updated automatically:

```{r, eval = FALSE}
run_id <- lin_run_start(ctx, ingest_id)
# … run the pipeline …
lin_run_complete(ctx, run_id, rows_produced = 42000L)

lin_stale(ctx)   # returns nodes whose upstream is newer than themselves
```

## Visualisation

```{r, eval = FALSE}
lin_plot(ctx)                             # uses visNetwork, ggraph, or igraph
lin_plot(ctx, highlight_node_id = s3_id) # highlight a node and its subtree
```
