---
title: "OpenLineage Integration: visualising R pipeline lineage in Marquez"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{OpenLineage Integration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Why OpenLineage?

[OpenLineage](https://openlineage.io) is the open standard for pipeline
lineage, adopted by Apache Airflow, Spark, dbt, Dagster, and many others.
By emitting OpenLineage events, `lineageR` lets R pipelines participate in
**the same lineage graph** as Python and SQL pipelines — without writing a
single line of extra glue code.

Benefits:

| Benefit | How |
|---|---|
| Free UI | [Marquez](https://marquezproject.ai) provides a graph browser, job history, and column lineage viewer |
| Interoperability | R, Python, dbt, Spark lineage in one namespace |
| Future-proof | Any tool that speaks OpenLineage (DataHub, Atlan, OpenMetadata) works automatically |

---

## Step 1 — Start Marquez with Docker

```{r marquez-setup}
lin_marquez_setup()
```

This prints the exact Docker command. Copy-paste it into a terminal:

```bash
docker run --name marquez \
  -p 5000:5000 -p 6000:6000 \
  marquezproject/marquez
```

Marquez API will be at `http://localhost:5000` and the UI at
`http://localhost:6000`.

---

## Step 2 — Configure lineageR to emit events

```{r configure}
library(lineageR)
library(DBI)
library(duckdb)

con <- dbConnect(duckdb(), "pipeline.duckdb")
ctx <- lineage_context(con)

# Enable OpenLineage emission — point at local Marquez
ctx <- lin_openlineage_config(ctx,
  url       = "http://localhost:5000/api/v1/lineage",
  namespace = "aviation-demo"
)
```

You can also use environment variables so the URL never appears in source
code:

```bash
export OL_URL="http://localhost:5000/api/v1/lineage"
export OL_NAMESPACE="aviation-demo"
export OL_ENABLED="true"
```

```{r configure-envvar}
# If env vars are set, just calling lineage_context() picks them up:
ctx <- lineage_context(con)   # OL_ENABLED=true, OL_URL and OL_NAMESPACE set
```

### NDJSON fallback

If `url` is not set (or if the HTTP call fails), events are appended to a
local NDJSON file. This is useful for offline development or audit trails:

```{r ndjson}
ctx <- lin_openlineage_config(ctx,
  ndjson_path = "pipeline_events.ndjson",
  enabled     = TRUE
)
```

---

## Step 3 — Register nodes and wire edges

This is identical to normal `lineageR` usage. Events are emitted
automatically when runs are recorded — no extra calls needed.

```{r register}
s3_id     <- lin_register_source(ctx, "s3_raw_flights",
               connection_string = "s3://my-bucket/raw/flights/")
ingest_id <- lin_register_transform(ctx, "ingest_flights",
               script_path = "pipelines/ingest.R", language = "r")
raw_id    <- lin_register_dataset(ctx, "flights_raw",
               table_name = "flights_raw")

lin_add_input(ctx,  ingest_id, s3_id)
lin_add_output(ctx, ingest_id, raw_id)
```

---

## Step 4 — Record a run (events are emitted automatically)

```{r run}
run_id <- lin_run_start(ctx, ingest_id)
# → emits a START event to Marquez

tryCatch({
  # ... run your pipeline code ...
  lin_run_complete(ctx, run_id, rows_produced = 48_231L)
  # → emits a COMPLETE event with input/output datasets
}, error = function(e) {
  lin_run_error(ctx, run_id, conditionMessage(e))
  # → emits a FAIL event with error details
})
```

### What the COMPLETE event looks like

```json
{
  "eventType": "COMPLETE",
  "eventTime": "2024-06-01T08:02:14Z",
  "run":  { "runId": "550e8400-e29b-41d4-a716-446655440000" },
  "job":  { "namespace": "aviation-demo", "name": "ingest_flights" },
  "inputs": [
    { "namespace": "aviation-demo", "name": "s3_raw_flights", "facets": {} }
  ],
  "outputs": [
    {
      "namespace": "aviation-demo",
      "name": "flights_raw",
      "facets": {
        "schema": {
          "_producer": "https://github.com/CathalByrneGit/lineageR",
          "fields": [
            { "name": "flight_id",  "type": "VARCHAR" },
            { "name": "origin",     "type": "VARCHAR" },
            { "name": "dep_time",   "type": "TIMESTAMP" }
          ]
        }
      }
    }
  ]
}
```

---

## Viewing lineage in Marquez

Open `http://localhost:6000` in a browser. After a few runs you will see:

- **Job graph** — a DAG of all transform steps in the `aviation-demo` namespace
- **Dataset lineage** — which datasets feed which transforms
- **Run history** — status (success / fail), run timestamps, row counts
- **Column lineage** — field-level provenance when `cell_expression_lineage`
  records exist

---

## Backfilling existing runs

If lineageR already has run history but Marquez was not configured yet, use
`lin_replay_events()` to backfill:

```{r replay}
lin_replay_events(ctx)
# or replay only recent runs:
lin_replay_events(ctx, since = as.POSIXct("2024-01-01"))
```

---

## Sharing a namespace with dbt

If a dbt project also pushes OpenLineage events to the same Marquez instance
and the same namespace, the full end-to-end lineage graph (dbt models →
lineageR transforms → ontologySpecR object types) will appear in a single
connected graph.

In `profiles.yml` (dbt-core ≥ 1.6 with the `dbt-openlineage` adapter):

```yaml
my_project:
  outputs:
    prod:
      type: duckdb
      openlineage_url: http://localhost:5000/api/v1/lineage
      openlineage_namespace: aviation-demo   # same namespace as lineageR
```

---

## Disabling emission

```{r disable}
ctx <- lin_openlineage_config(ctx, enabled = FALSE)
```

Or set `OL_ENABLED=false` in the environment before starting R.
