---
title: "targets Integration: lin_from_targets() walkthrough"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{targets Integration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

`lin_from_targets()` reads a **targets** pipeline manifest and registers each
target as a `transform` node, wiring dependency edges automatically.

This lets you track lineage without manually calling `lin_register_transform()`
and `lin_add_input()` / `lin_add_output()` for every step.

## Example pipeline (`_targets.R`)

```{r, eval = FALSE}
library(targets)

list(
  tar_target(raw_data,     read_csv("data/raw.csv")),
  tar_target(clean_data,   clean(raw_data)),
  tar_target(model_output, fit_model(clean_data)),
  tar_target(report,       render_report(model_output))
)
```

## Importing the pipeline into lineageR

```{r, eval = FALSE}
library(lineageR)
library(duckdb)

con <- DBI::dbConnect(duckdb::duckdb(), ":memory:")
ctx <- lineage_context(con)

node_ids <- lin_from_targets(ctx, pipeline_path = "_targets.R")
```

`lin_from_targets()` returns a named list mapping each target name to its
`node_id`:

```
$raw_data
[1] "xxxxxxxx-xxxx-4xxx-xxxx-xxxxxxxxxxxx"

$clean_data
[1] "yyyyyyyy-yyyy-4yyy-yyyy-yyyyyyyyyyyy"
...
```

## What gets created

For each target a `transform` node is created.  Dependencies between targets
become `"produces"` edges:

```
raw_data  ──produces──>  clean_data  ──produces──>  model_output  ──produces──>  report
```

## Extending with source and dataset nodes

After importing the targets manifest you can graft in source and dataset nodes
manually:

```{r, eval = FALSE}
src_id    <- lin_register_source(ctx, "raw_csv", connection_string = "data/raw.csv")
report_ds <- lin_register_dataset(ctx, "report_html", table_name = "output/report.html")

lin_add_input(ctx,  node_ids[["raw_data"]], src_id)       # csv -> raw_data
lin_add_output(ctx, node_ids[["report"]],   report_ds)    # report -> HTML file

lin_plot(ctx)
```

## Provenance after pipeline run

```{r, eval = FALSE}
run_id <- lin_run_start(ctx, node_ids[["model_output"]])
# … fit model …
lin_run_complete(ctx, run_id, rows_produced = 1L)

lin_stale(ctx)     # downstream of model_output now has a fresh timestamp
lin_impact(ctx, node_ids[["raw_data"]])   # everything downstream of raw_data
```
